### 2017-06-01 distr-03-design.tex
Refactor distr so there is a design section again

% intro parallelisation, distribution

This section describes how we created a distributed version of the C++
algorithm. We didn't use the advanced intra-node parallelization techniques
described in the previous section, but found that straightforward
multi-threaded parallelization techniques sufficed: the distributed performance
is dominated by (remote) memory latency, not by node parallelism. As our
analysis will show, our pipelining optimizations will fully overlap the
parallel calculations with (pre)loading the data from remote hosts.
The details of the multi-threaded parallelism
will be discussed only where appropriate. The
distributed design follows a master-worker paradigm, where
the master controls the sequence of parallel operations and the workers perform the
calculations. For thread parallelism, we annotate the program with
OpenMP~\cite{OpenMPSpec}. For distributed communication, we use
MPI~\cite{Forum:1994:MMI:898758}.

@@

\subsection{Data distribution}

@@

\subsection{RDMA Distributed Key-Value Store}

@@

The distributed algorithm implementation stores $\pi$ in a DKV store in the
collective memory of the
cluster, where $\pi[i]+\sum\phi[i]$ is the value for key $i$.
%
We decided not
to use any of the existing implementations of RDMA DKV stores, and instead
build our own DKV store on top of the Infiniband ib-verbs~API.
%
This is motivated by the fact that our use case is unique in a number
of important aspects. First, the number of keys and values to be stored are
constant throughout the execution of the algorithm. For instance, no records
are dynamically inserted or deleted from the DKV after their initial insertion,
although the value associated with each key can be updated. Additionally, all
keys are of the same type and represent a unique vertex in the graph.
Therefore, keys can be represented as integer identifiers. Moreover, each key
points to a vector of $K+1$ floats as its value.
%
This data representation can be exploited to simplify the management of
read and write requests. Similarly, the keys can be deterministically
partitioned across the cluster using a fast hash function.

The algorithm's access pattern is well controlled because the
computation is partitioned into stages that are separated by barriers. A
single stage either reads values or updates values. The algorithm ensures that updates
are always targeting unique elements,
therefore, there are no read/write or write/write hazards. The existing DKV
stores (RamCloud~\cite{Ousterhout:2015:RSS:2818727.2806887}, FaRM~\cite{179767},
and Herd~\cite{Kalia:2014:URE:2740070.2626299}) incur
overhead for insert/delete flexibility, concurrency control, or variable-sized
values. Our use case allows us to do any operation in exactly one RDMA read
or RDMA write.
%
Herd argues that replacing RDMA reads by
an RPC to the server, followed by an unreliable write, is faster.
%
However,
the paper shows this only holds for small payload packets, up to 256B. A $\pi$
packet in our application is typically thousands to hundreds of thousands
of 4-byte floats.

@@

\subsection{Implementation of distributed parallelism}

This section describes the parallelization of each of the stages of the
algorithm's main loop. All but the first stage justify parallelization for
high values of combinations of the parameters mini-batch size $|\Minibatch|$,
number of vertices in the mini-batch $M$,
number of communities $K$, neighbor sample size $|\Neighbors|$, and held-out set
size $|\cE_h|$.

The first stage, mini-batch selection (line 3 in Algorithm~\ref{algorithm}),
is done by the master and it is
not itself parallelized. However, the distributed implementation overlaps
its execution using pipeline parallelism while the workers are
performing \textit{update\_phi}. The mini-batch is partitioned equally over
the workers and the relevant sections of \Edges are distributed together with the
mini-batch subsets.

@@

%\looseness=-1
These updates are used by \textit{update\_pi} (line 9 in
Algorithm~\ref{algorithm}) locally, besides the value of
$\pi/\phi$ for the mini-batch vertices. The update to $\pi$ requires
$M \times K$ operations, and it is done in parallel over mini-batch vertices. Because of
memory consistency, this stage awaits completion of \textit{update\_phi} with
an MPI barrier. After the calculation, the updated values of $\pi,~\sum\phi$
are written to the DKV store.

\textit{update\_beta} (lines 12 and 15 in Algorithm~\ref{algorithm}) requires $|\Minibatch| \times K$ operations. It is preceded by an MPI
barrier to ensure that up-to-date $\pi$ values are read. \textit{update\_beta}
is split into four steps. The first stage partitions the mini-batch across
machines to calculate contributions to $g_{ab}(\theta)$. The values of $\pi$ for
the local mini-batch vertices are loaded from the DKV store. These calculations
are done with thread parallelism over the mini-batch vertices. In the second step,
a multi-threaded summation is performed to calculate the contributions per
machine, followed by a distributed summation using an MPI reduce operation. The
third step calculates $\beta$ from $\theta$ sequentially at the master. This
operation takes only $K$ steps. The resulting $\beta$ is broadcast with MPI
to the workers.

@@

\subsection{Pipelining of computation and loading $\pi$}

>>>>

\subsection{Design of distributed implementation}
% intro parallelisation, distribution

This section describes how we create and evaluate a distributed version of the C++
algorithm. We didn't use the advanced intra-node parallelization techniques
described in the previous section, but found that straightforward
multi-threaded parallelization techniques sufficed: the distributed performance
is dominated by (remote) memory latency, not by node parallelism. As our
analysis will show, our pipelining optimizations will fully overlap the
parallel calculations with (pre)loading the data from remote hosts.
The details of the multi-threaded parallelism
will be discussed only where appropriate. The
distributed design follows a master-worker paradigm, where
the master controls the sequence of parallel operations and the workers perform the
calculations. For thread parallelism, we annotate the program with
OpenMP~\cite{OpenMPSpec}. For distributed communication, we use
MPI~\cite{Forum:1994:MMI:898758}.

@@

\subsubsection{Data distribution}

@@

\subsubsection{RDMA Distributed Key-Value Store}

@@

The distributed algorithm implementation stores $\pi$ in a DKV store in the
collective memory of the
cluster, where $\pi[i]+\sum\phi[i]$ is the value for key $i$.
%
We decided not
to use any of the existing implementations of RDMA DKV stores, and instead
build our own DKV store on top of the Infiniband ib-verbs~API.
%
This is motivated by the fact that our use case is unique in a number
of important aspects. First, the number of keys and values to be stored are
constant throughout the execution of the algorithm: no records
are dynamically inserted or deleted from the DKV store after their initial insertion,
although the value for any key can be updated. Additionally, all
keys are of the same type and represent a unique vertex in the graph
so, they can be represented as integer identifiers. Each key
has a vector of $K+1$ floats as its value.
%
This data representation can be exploited to simplify the management of
read and write requests. Similarly, the keys can be deterministically
partitioned across the cluster using a fast hash function.

The algorithm's access pattern is well controlled because the
computation is partitioned into stages that are separated by barriers. A
single stage either reads values or updates values, and the algorithm ensures that updates
are always targeting unique elements.
Therefore, there are no read/write or write/write hazards. The existing DKV
stores (RamCloud~\cite{Ousterhout:2015:RSS:2818727.2806887}, FaRM~\cite{179767},
and Herd~\cite{Kalia:2014:URE:2740070.2626299}) incur
overhead for insert/delete flexibility, concurrency control, or variable-sized
values. Our use case allows us to do any operation in exactly one RDMA read
or RDMA write.
%
Herd argues that replacing RDMA reads by
an RPC to the server, followed by an unreliable write, is faster.
%
However,
the paper shows this only holds for small payload packets, up to 256B. A $\pi$
packet in our application is typically thousands to hundreds of thousands
of 4-byte floats.

@@

\subsubsection{Implementation of distributed parallelism}

This section describes the parallelization of the stages of the
algorithm's main loop. All but the first stage,
mini-batch selection (line 3 in Algorithm~\ref{algorithm}),
justify parallelization for
high values of combinations of the parameters mini-batch size $|\Minibatch|$,
number of vertices in the mini-batch $M$,
number of communities $K$, and sizes of neighbor sample set~$|\Neighbors|$ and held-out
set~$|\cE_h|$.

Mini-batch selection is done by the master and it is
not itself parallelized. However, the distributed implementation overlaps
its execution using pipeline parallelism while the workers are
performing \textit{update\_phi}. The mini-batch is partitioned equally over
the workers and the relevant sections of \Edges are distributed together with the
mini-batch subsets.

@@

%\looseness=-1
These updates are used by \textit{update\_pi} (line 9 in
Algorithm~\ref{algorithm}) locally, besides the value of
$\pi/\phi$ for the mini-batch vertices. The update to $\pi$ requires
$M \times K$ operations, and it is done in parallel over mini-batch vertices. For
memory consistency, this stage awaits completion of \textit{update\_phi} with
an MPI barrier. After the calculation, the updated values of $\pi,~\sum\phi$
are written to the DKV store.

\textit{update\_beta} (lines 12 and 15 in Algorithm~\ref{algorithm}) requires $|\Minibatch| \times K$ operations. It is preceded by an MPI
barrier to ensure that up-to-date $\pi$ values are read. \textit{update\_beta}
is split into four steps. The first step partitions the mini-batch across
machines to calculate contributions to $g_{ab}(\theta)$. The values of $\pi$ for
the local mini-batch vertices are loaded from the DKV store. These calculations
are done with thread parallelism over the mini-batch vertices. In the second step,
a multi-threaded summation is performed to calculate the contributions per
machine, followed by a distributed summation using an MPI reduce operation. The
third step calculates $\beta$ from $\theta$ sequentially at the master. This
operation takes only $K$ steps. The resulting $\beta$ is broadcast with MPI
to the workers.

@@

\subsubsection{Pipelining of computation and loading $\pi$}

### 2017-06-01 distr-04-evaluation.tex
Tiny textual fixes
Refactor distr so there is a design section again

All of the experiments reported in this section utilized
publicly available graphs from the Stanford Large Network Data set
Collection~(SNAP) \cite{snapnets}. Table~\ref{table-snap} lists the
collection of data sets used.

@@

In order to evaluate the horizontal scalability of the distributed
implementation we tested the system's performance across different cluster
sizes while holding the problem size constant. For this study, we used the
com-Friendster graph as it contains the largest number of vertices and edges.
%
Figure~\ref{fig-strong-scaling}-a presents the execution time of 2048 algorithm
iterations across multiple cluster sizes. The x-axis starts from 8 worker nodes
as the data set is too large to fit into the collective memory of a smaller
cluster.
%
As shown in the figure, the execution
time steadily decreases by increasing the cluster size.
A deeper analysis of the individual execution phases of the algorithm provides
insights into the scalability of its building blocks.
%
In addition to the total execution time, Figure~\ref{fig-strong-scaling}-a
presents the cumulative time spent in individual computational phases across
iterations. Moreover, Figure~\ref{fig-strong-scaling}-b presents the speedup
achieved for the same experiments reported in
Figure~\ref{fig-strong-scaling}-a. As is clearly shown, the dominant phase of
the execution is \textit{update\_phi\_pi}.
The reported total time for each cluster size is significantly less than the
sum of the execution times of the individual phases. This is due to the
overlapping execution of the two most expensive phases, namely,
\textit{update\_phi\_pi} and
mini-batch deployment. Both of these phases initially gain significant speedup with the
addition of compute nodes. However, the speedup curve gradually slows down for
larger cluster sizes as the work granularity of each worker node decreases,
limiting their resource utilization.
%
The time spent in \textit{update\_beta\_theta} remains relatively constant across cluster sizes
as it performs an insignificant amount of work compared to the synchronization
overhead of a collective MPI reduction operation contained within it.

@@

Figure~\ref{fig-pipeline} presents the execution time of 1024 algorithm
iterations on a 64-node cluster with the pipelines enabled and disabled.
Naturally, increasing the number of communities causes a proportional increase
in execution time. However, when pipelining is enabled, some of the
incurred network latency is hidden by overlapping it with computation.
Moreover, since both computation time and network latency increase with larger
$K$, the benefit of pipelining increases. This can be observed through the
widening gap between both lines depicted in Figure~\ref{fig-pipeline}.
Large cluster
configurations exhibit higher bandwidth demands and are more sensitive
to network latency:
given that $\pi$ accesses are random, a node in a cluster of size $C$ must
fetch $(C-1)/C$ of all read requests over the network.

>>>>

All of the experiments reported in this section utilized
publicly available graphs from the SNAP collection, see Table~\ref{table-snap}.

@@

In order to evaluate the horizontal scalability of the distributed
implementation we tested the system's performance across different cluster
sizes while holding the problem size constant. For this study, we used the
com-Friendster graph as it contains the largest number of vertices and edges.
%
Figure~\ref{fig-strong-scaling}-a presents the execution time of 2048 algorithm
iterations across multiple cluster sizes. The x-axis starts from 8 worker nodes
as the data set is too large to fit into the collective memory of a smaller
cluster.
%
As shown in the figure, the execution
time steadily decreases by increasing the cluster size.
A deeper analysis of the individual execution phases of the algorithm provides
insights into the scalability of its building blocks.
%
In addition to the total execution time, \mbox{Figure~\ref{fig-strong-scaling}(a)}
presents the cumulative time spent in individual computational phases across
iterations. Moreover, \mbox{Figure~\ref{fig-strong-scaling}(b)} presents the speedup
achieved for the same experiments reported in
Figure~\ref{fig-strong-scaling}(a). As is clearly shown, the dominant phase of
the execution is \textit{update\_phi\_pi}.
The reported total time for each cluster size is significantly less than the
sum of the execution times of the individual phases. This is due to the
overlapping execution of the two most expensive phases, namely,
\textit{update\_phi\_pi} and
mini-batch deployment. Both of these phases initially gain significant speedup with the
addition of compute nodes. However, the speedup curve gradually slows down for
larger cluster sizes as the work granularity of each worker node decreases,
limiting their resource utilization.
%
The time spent in \textit{update\_beta\_theta} remains relatively constant across cluster sizes
as it performs an insignificant amount of work compared to the synchronization
overhead of a collective MPI reduction operation contained within it.

@@

Figure~\ref{fig-pipeline} presents the execution time of 1024 algorithm
iterations on a 64-node cluster with the pipelines enabled and disabled.
Naturally, increasing the number of communities causes a proportional increase
in execution time. However, when pipelining is enabled, the
incurred network latency is partly hidden by overlapping it with computation.
Moreover, since both computation time and network latency increase with larger
$K$, the benefit of pipelining increases. This can be observed through the
widening gap between both lines depicted in Figure~\ref{fig-pipeline}.
Large cluster
configurations exhibit higher bandwidth demands and are more sensitive
to network latency:
given that $\pi$ accesses are random, a node in a cluster of size $C$ must
fetch $(C-1)/C$ of all read requests over the network.
